---
title: "Intro to Network Analysis with R"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r load_packages}
#install.packages("igraph")
#install.packages("igraphdata")
#install.packages("ape")
library(igraphdata)
library(igraph)
library(ape)

```

Load data

```{r}
data(UKfaculty)
```

Some graphics setup

``` {r}

#default margins
def_marg <- c(5.1, 4.1, 4.1, 2.1)

#no margins
no_marg <- c(0, 0, 0, 0)

```

Plot of our network 

```{r}

par(mar=no_marg)
plot(UKfaculty, 
     vertex.size = 9,
     vertex.color = "dodgerblue",
     vertex.frame.color= "dodgerblue",
     vertex.label.family = "Helvetica", 
     vertex.label.cex = 0.5,
     vertex.label.color = "white",
     edge.arrow.size = 0.2)
```

# Exploring our Network

## Overview

What are the most general characteristics of this network?

```{r}
UKfaculty
```

Does the graph has multiple edges per pair of nodes? Does the graph have any loops?

```{r}
is.simple(UKfaculty)
```

# Now you try it!

How many nodes does the karate network have? Is the karate network directed or indirected? What vertex and edge attributes does the network have?

```{r}
data(karate)
karate
```

# Taking a closer look: Network Attributes 

## Graph Attributes

```{r}
graph.attributes(UKfaculty)
```

## Vertex Attributes

```{r}
vertex.attributes(UKfaculty)

V(UKfaculty)[[1:5]]

V(UKfaculty)$Group

table(V(UKfaculty)$Group)
```

## Edge Attributes

```{r}
edge.attributes(UKfaculty)

E(UKfaculty)[[1:5]]

E(UKfaculty)$weight

summary(E(UKfaculty)$weight)

par(mar=c(2,2,2,2))
hist(E(UKfaculty)$weight)
```

# Accessing and Exploring Specific Elements 

## Exploring vertices 

All vertices

```{r}
V(UKfaculty)
```

Specific node connections 

Which nodes are connected to node 1?

```{r}

# outgoing ties
neighbors(UKfaculty, 1, mode = "out")

#incoming ties
neighbors(UKfaculty, 1, mode = "in")

# all ties
neighbors(UKfaculty, 1, mode = "all")

```

Which nodes have less than 10 connections?

```{r}
V(UKfaculty)[degree(UKfaculty) < 10]
```

## Exploring edges

Does a given edge exist?

```{r}

UKfaculty[1,2]
UKfaculty[1,44]
UKfaculty[44,1]

```

Edges between specific nodes

```{r}
E(UKfaculty)[1 %--% 2]

# regardless of the direction
E(UKfaculty)[1 %--% 44]

# from 44 to 1
E(UKfaculty)[1 %<-% 44]

# from 1 to 44
E(UKfaculty)[1 %->% 44]
```

Edges containing specific vertices

```{r}

# All incoming edges of a vertex
E(UKfaculty)[to(1)]

# All outgoing edges of a vertex
E(UKfaculty)[from(1)]

```

# Now you try it!

Let's take a closer look at the karate network! 

Running this code will tell us the general characteristics of the network and the name of its attributes. 

```{r}
karate
```

Replace every [xxx] in the following code to answer each question. 

How many factions does the karate network have? How many nodes belong to each faction?

```{r}
table(V(karate)$Faction)
```


```{r}
V(karate)[[1:5]]
```

# Network Visualization 

Basic plot

```{r}
#setting wider margins
par(mar=c(0,0,0,0))

#basic plot
plot(UKfaculty)
```

Option 1: Setting visual properties outside the plot function

We add attributes to the network

Step 1 Set vertex and/or edge attributes

```{r}

# vertex properties 
V(UKfaculty)$color <- "dodgerblue"

# vertex label properties
V(UKfaculty)$label.color <- "white"
V(UKfaculty)$label.cex <- 0.5

# edge properties
E(UKfaculty)$color <- "gray"

```

Step 2 Plot network

```{r}

plot(UKfaculty)

```

Option 2 Setting visual properties inside the plot function

We use the function arguments

```{r}

plot(UKfaculty, 
     # === vertex properties 
     vertex.color = "dodgerblue",
     vertex.frame.color = "dodgerblue",
     vertex.size = 9,
     # === vertex label properties
     vertex.label.cex = 0.5,
     vertex.label.color = "white",
     vertex.label.family = "Helvetica",
     # === edge properties
     edge.color = "gray",
     edge.width = 0.5,
     edge.arrow.size = 0.2)
```


This help us add different visual properties to subsets of edges or nodes. 

Coloring nodes by group

```{r}
table(V(UKfaculty)$Group)

mycolors <- c("#55C667FF",
              "#404788FF",
              "#481567FF",
              "#33638DFF")

V(UKfaculty)$color <- mycolors[V(UKfaculty)$Group]
V(UKfaculty)$frame.color <- mycolors[V(UKfaculty)$Group]

plot(UKfaculty)
```

We can also combine both options. This is usually what we do. 

```{r}
plot(UKfaculty,
     vertex.size = 9,
     # === vertex label properties
     vertex.label.family = "Helvetica",
     # === edge properties
     edge.width = 0.5,
     edge.arrow.size = 0.1)
```


Coloring specifc edges

```{r}
E(UKfaculty)$color <- "gray"
E(UKfaculty)$color[E(UKfaculty)[47%--%28]] <- "red"

plot(UKfaculty, 
     # === vertex properties 
     vertex.size = 9,
     # === vertex label properties
     vertex.label.cex = 0.5,
     vertex.label.family = "Helvetica",
     # === edge properties
     edge.width = 0.5,
     edge.arrow.size = 0.1)

```

# Now you try it

Create your own plot of the karate data. Try changing different parameters. 

```{r}

```

# Descriptive Network Statistics

```{r}
# Density 
graph.density(UKfaculty)

# Reciprocity 
reciprocity(UKfaculty)

# Dyad Census 
dyad_census(UKfaculty)

# Transitivity 
transitivity(UKfaculty)
transitivity(UKfaculty, type="local")
triad_census(UKfaculty)

# Reachability 
average.path.length(UKfaculty)

diameter(UKfaculty, weights = NA)
diameter(UKfaculty, directed = FALSE, weights = NA)

farthest_vertices(UKfaculty)

# Connected components
is.connected(UKfaculty, mode = "strong")

# Articulation points
articulation.points(UKfaculty)

```

# Centrality 

## Degree Centrality 

```{r}

# Indegree 
degree(UKfaculty, mode="in")
par(mar= def_marg)
hist(degree(UKfaculty, mode = "in"),
     breaks=1:vcount(UKfaculty)-1, 
     main="Indegree histogram")

# Outdegree
degree(UKfaculty, mode="out")
hist(degree(UKfaculty, mode = "out"),
     breaks=1:vcount(UKfaculty)-1, 
     main="Outdegree histogram")

# Combine indegree and outdegree 

deg <- degree(UKfaculty, mode="all")

par(mar= no_marg)

plot(UKfaculty, 
     vertex.size=deg)

plot(UKfaculty, 
     vertex.size=sqrt(deg)+1,
     vertex.label.cex = 0.4,
     edge.arrow.size = 0.1)
```

## Closeness Centrality 

```{r}
closeness(UKfaculty, mode="all", weights=NA) 

centr_clo(UKfaculty, mode="all", normalized=T)

```

## Betweenness Centrality 

```{r}
betweenness(net, directed=T, weights=NA)

edge_betweenness(net, directed=T, weights=NA)

centr_betw(net, directed=T, normalized=T)
```

# Community Detection

Data preparation 

```{r}
# convert to undirected graph
und_UKfaculty <- as.undirected(UKfaculty)
```

Conduct community structure procedure


```{r}
kc <- fastgreedy.community(und_UKfaculty)
```

Explore results 

```{r}
class(kc)

length(kc)

sizes(kc)

modularity(kc)
```

Plot results 
```{r}
plot(kc, und_UKfaculty)
dendPlot(kc, mode = "phylo")
```

# Now you try it!

Conduct a community structure procedure with 
the karate data. How many groups do you obtain?

```{r}

```

